---
title: "EPI 569 -- Exercise 1: Natural history of infectious diseases"
author: "Ben Lopman, Elizabeth Rogawski McQuade"
output: html_document
widescreen: true
smaller: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(tidyverse)
library(fitdistrplus)
knitr::opts_chunk$set(fig.height = 4, fig.width = 4, warning = FALSE)
```

## Background

- In December 2009, an outbreak occurred following a restaurant exposure to a foodborne pathogen over many days.  More information is in the paper reporting the outbreak, if you are interested, but not necessary that you read the article. https://pubmed.ncbi.nlm.nih.gov/21524343/  

- CDC and the local health departments conducted a study of those who fell ill, as well as their household contacts.  They found that this outbreak was a result of oysters consumed at the restaurant that were contaminated with norovirus. 

- This dataset presents a good opportunity to study natural history, because there is a *known time of exposure* (in hours) for the cases who dined at the restaurant. In addition, there are data on illnesses among the household contacts who did not dine at the restaurant.  That way, we can also look at the natural history among household contacts.   

### The spreadsheet ‘NaturalHistoryExercise.csv’ contains the data, which you can save from Canvas or import directly into R from the web as follows:

```{r, include=TRUE, eval=TRUE}
data <- read.csv(url("https://raw.githubusercontent.com/blopman/epi569/master/NaturalHistoryExercise.csv")) #%>% rename(HOUSEHOLD_ = ï..HOUSEHOLD_) #you may need to run the commented out code if the HOUSEHOLD_ variable is read in wrong
```

- You can see all the data on times of exposure, onset of symptoms and time of last symptom episode. The **HOUSEHOLD_** variable indicated which household the person was a member; and **HOUSEHOLD_INDEX** indicated if the person got ill at the restaurant. The other fields should be self-explanatory.

- We have calculated the key values for each person for you.  For each, there is a value for the linear scale (DAYS) and log scale (LOG DAYS).  You can explore the rest of the dataset, but most wont be needed for this exercise.
   + **INDEX_INCUBATION_DAYS**	-- The time from dining at the restaurant to onset of symptoms in index cases
   + **SERIAL_INTERVAL_DAYS**	-- The time from onset of illness in the index case to the onset in the household contact
   + **DURATION_DAYS** -- The time from the onset of symptoms to the end of illness in all cases.

## It is recommended that you use **R**

- If you use **R**, the following functions will be useful to you:
   + median(data$**VARIABLE_NAME**, na.rm=TRUE)    -- Median
   + sd(data$**VARIABLE_NAME**, na.rm=TRUE)        -- Standard deviation
   + log(data$**VARIABLE_NAME**, na.rm=TRUE)       -- Natural log
   + exp(data$**VARIABLE_NAME**, na.rm=TRUE)       -- Exponentiate

## Now for your work.  For this outbreak:

### Question 1 **(1pt)**
- Calculate *the incubation period*:
  + the median of the incubation period
  + the median of the log incubation period
  + the standard deviation of the log incubation period
  + the dispersion, which is exp(sd)

### ANSWER

    
- Because incubation periods are log-normally distributed, 66% of cases should fall within the median/exp(sd) and median*exp(sd). Does that appear to be the case?

### ANSWER  


- Now, *use the following commands* to plot the distribution of this variable and a fitted log-normal distribution as follows:
  + Create a variable without all the missing data
  + Fit a log normal distribution
  + Look at the parameters of this log normal distribution 
  + Plot the log normal distribution along with the data (histogram)
      
```{r, eval = TRUE, echo=TRUE}
incubation<-data$INDEX_INCUBATION_DAYS[!is.na(data$INDEX_INCUBATION_DAYS)]
incubation_fit_ln <- fitdist(incubation, "lnorm")
summary(incubation_fit_ln)
denscomp(incubation_fit_ln)
```

### Question 2 **(1pt)**
- Calculate the same four values (median of the incubation period, the median of the log incubation period, the standard deviation of the log incubation period, and the dispersion) and generate the same plot, this time for *the serial interval*.
  + Did you find a wide distribution of serial intervals (i.e. observed serial intervals of many different lengths)?  If so, why do think that could be?


### ANSWER

      
### Question 3 **(1pt)**
- Finally, calculate the same four values (median of the incubation period, the median of the log incubation period, the standard deviation of the log incubation period, and the dispersion) and generate the same plot, this time for *the duration of illness*.

### ANSWER


### Question 4 **(1pt)**
- Are the values you calculated for (1-3) consistent with norovirus as the cause of the outbreak?  Research online for the natural history parameters for norovirus to justify your answer.

### ANSWER

    
### Question 5 **(1pt)**
- Calculate the secondary attack rate among household contacts.
  + *(number ill /number exposed)*

### ANSWER


### Question 6 [Extra Credit] **(0.5pt)**
- Calculate and plot the distribution of secondary attack rates by household.

### ANSWER

