---
title: "EPI 569 -- Exercise 3: Diagnostics and Seroepidemiology"
author: "Victoria Ngo"
format: html
editor: visual
widescreen: true
smaller: true
self-contained: true
---

```{r setup, include=FALSE, echo = FALSE, eval = TRUE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(kableExtra)
library(ggplot2)
```

## Introduction

### The datafile *India_RotaSerology_updated.csv* contains measurements on antibodies against rotavirus for children from birth to 18 years of age and measurements of virus detection at 36 months.

```{r}
sero <- data.frame(read.csv("https://raw.githubusercontent.com/blopman/epi569/master/India_RotaSerology_updated.csv"))
view(sero)
```

-   You have data on about 350 children.

-   Blood was drawn at birth and at 36-monthly intervals (every 3 years)

    -   Serum IgG and IgA was measured by EIA using serial dilutions.
    -   Dilutions were 2-fold, with dilutions ranging from 1:10, 1:20; 1:40 to a maximum of 1:2560.

-   Additionally, at 36 months, both an enzyme immunoassay (EIA) test and a polymerase chain reaction (PCR) test were used to detect rotavirus in the children's stool.

    -   EIA results were recorded as positive (virus detected) or negative (virus not detected).
    -   PCR results were recorded as cycle thresholds, i.e. the number of PCR cycles needed to detect rotavirus. Lower cycle thresholds indicate higher viral loads. The maximum number of cycles run was 40. If fewer than 40 cycles were recorded, rotavirus was detected by PCR; if 40 cycles were recorded, rotavirus was not detected.

### Question 1 **(3pts)**

-   What is the correct summary measure of antibody concentrations? State the value should be analyzed and the appropriate summary statistic. (Hint: see the lecture slides on "Quantifying antibodies")

### ANSWER

::: {.callout-note appearance="minimal" icon="false"}
The correct summary measure of antibody concentrations is the geometric mean of the log of the reciprocal titer values. The values should be analyzed on a log scale of the reciprocal dilution values, specifically log base 10. The appropriate summary statistic is the geometric mean. This is calculated by taking the antilog ($10^x$) of the arithmetic mean of the log-transformed reciprocal dilutions.
:::

-   Calculate this summary measure for IgG and IgA by age.

### ANSWER

```{r, echo = TRUE}
# here's some code to summarise IgG by age
igg_gmt <- sero %>% 
  group_by(Age_Month) %>% 
  summarize(geo_mean_igg = 10^(mean(log10(IgG))))
print(igg_gmt)

iga_gmt <- sero %>% 
  group_by(Age_Month) %>% 
  summarize(geo_mean_iga = 10^(mean(log10(IgA))))
print(iga_gmt)


# edit code above (or write new code) to also summarize IgA by age
```

-   Describe in words the patterns of the IgG and IgA summary statistic over age.

### ANSWER

::: {.callout-note appearance="minimal" icon="false"}
**IgG**

At birth, newborns have a high IgG geometric mean, suggesting that newborns have elevated IgG levels due to maternal antibody transfer. The IgG values show an initial decline in early childhood. This is followed by a steady increase with age.

**IgA**

At birth, newborns have a very low IgA geometric mean that shows a continuous increase with age. After birth, there is a pronounced IgA growth during adolescence.
:::

-   Which antibody (IgG or IgA) would you use to estimate incidence (new infections)?

### ANSWER

::: {.callout-note appearance="minimal" icon="false"}
IgA would be better for estimating the incidence of new infections. In comparison to IgG, IgA has an earlier spike of markers after symptom onset. Additionally, IgA is typically involved with mucosal immunity and is the marker of recent immune response in the respiratory and gastrointestinal tracts where many infections occur.
:::

### Question 2 **(2pts)**

-   Next, convert the antibody data into two binary variables representing:
    -   Protective antibodies: An IgA titer of \>= 90 units is considered to be protective against future disease.
    -   Ever infected: An IgA titer of \>= 20 units indicates having ever been infected with rotavirus.
    -   Note: for ease of analysis, the variables can be coded as 1 = positive (i.e. above threshold as defined above); 0 = negative (i.e. below threshold as defined above)

```{r, echo = TRUE}
#binary variable for Protective Antibodies
sero$Protective_Antibodies <- ifelse(sero$IgA >= 90, 1, 0)

#binary variable for Ever Infected
sero$Ever_Infected <- ifelse(sero$IgA >= 20, 1, 0)
```

-   What proportion of children have protective antibody levels at each age in months (0, 36,... 216m)?
-   What proportion of children at each age have ever had a rotavirus infection?

### ANSWER

```{r}
# we suggest reporting the proportions requested above by each age group in a table format
proportions_by_age <- sero %>%
  group_by(Age_Month) %>%
  summarise(
    proportion_protective = round(mean(Protective_Antibodies, na.rm = TRUE), 2),
    proportion_ever_infected = round(mean(Ever_Infected, na.rm = TRUE), 2)
  )
proportions_by_age
```

### Question 3 **(2pts)**

-   Calculate R0, the basic reproduction number. Assume life expectancy is 68 years, average age of 1st infection is 5.5 years, and Type II mortality. See the lecture slides on calculating Ro from the average age of first infection in the "Analysis and modeling of serology data" lecture

### ANSWER

```{r}
life_expectancy <- 68
average_age_infection <- 5.5

R0 <- 1 + round(life_expectancy / average_age_infection, 2)

R0
```

-   What is the herd immunity threshold for rotavirus in this population? (Hint: You may need to look back to previous lectures on Ro and the HIT.)

### ANSWER

```{r}
HIT <- 1 - (1 / R0)
HIT
```

-   Do you think that this is a reasonable level of coverage to be achieved? Put another way, can rotavirus be eradicated? Explain your reasoning.

### ANSWER

::: {.callout-note appearance="minimal" icon="false"}
I think achieving a 92.51% HIT and eradicating rotovirus would be ambitious and challenging. While the rotovirus vaccine efficacy is extremely effective and high vaccination rates in infants and young children can lead to herd immunity, a poor public health infrastructure and vaccine hesitancy in endemic countries creates barriers to achieving herd immunity.
:::

### Question 4 **(3pts)**

-   At 36 months of age, calculate the mean numbers of cycle thresholds for PCR detection of rotavirus in this sample separately by EIA test status (i.e., the mean for those who test EIA negative and those who test EIA positive).

### ANSWER

```{r}
# HINT: similar to Q1, you will need the following functions: filter(), group_by(), summarise()

mean_pcr_ct <- sero %>%
  filter(Age_Month == 36) %>%
  group_by(EIA) %>%
  summarize(mean_PCR_Ct = round(mean(PCR_Ct, na.rm = TRUE),2))

print(mean_pcr_ct)
```

-   Additionally, calculate the percentage of children who tested positive by EIA and by PCR (e.g., rotavirus detected at less than 40 cycles), respectively. Compare these two values. Explain any differences you see.

### ANSWER

```{r}
# HINT: you will need to dichotomize PCR results into positive or negative

# Calculating percentage of children who tested positive by EIA, accounting for NAs
eia_positive_percentage <- sero %>%
  summarise(percentage_EIA_positive = sum(EIA == "Positive", na.rm = TRUE) / sum(!is.na(EIA)) * 100) 

# Create binary variable for PCR Status: 1 for positive (Ct < 40), 0 for negative (Ct >= 40)
sero$PCR_Status <- ifelse(sero$PCR_Ct < 40, 1, 0)

# Calculate percentage of children who tested positive by PCR (Ct < 40)
pcr_positive_percentage <- sero %>%
  summarise(percentage_PCR_positive = sum(PCR_Status == 1, na.rm = TRUE) / sum(!is.na(PCR_Status)) * 100)

# Combine the results into a single data frame for comparison
comparison <- data.frame(
  Measure = c("EIA Positive", "PCR Positive"),
  Percentage = c(eia_positive_percentage$percentage_EIA_positive, 
                 pcr_positive_percentage$percentage_PCR_positive)
)
comparison


```

-   Describe the differences between what is being detected by IgG/IgA levels and what is being detected by EIA/PCR.

### ANSWER

::: {.callout-note appearance="minimal" icon="false"}
-   The IgG/IgA reflects the immune system's long-term response and the immunity to the infection, whereas the EIA/PCR reflects both the current infection of a pathogen and the historical responses due to testing viral genetic material and antibody response.
:::
